Browsernative Microservices

### 2017-02-07 paper.tex
pdf/tex build

Computers can scale, humans can't. Ever since a program or complex
system made by humans has been constrained by humans mental
capabilities. Like in the analogy of juggling balls, our brain can just
``juggle'' a few things at a time. Rich Hickey, the inventor of the
programming language \emph{Clojure} gave an inspirational keynote on the
topic of \textbf{simplicity}.\footnote{\href{https://www.youtube.com/watch?v=rI8tNMsozo0\&t=46s}{Rails
  Conf 2012 Keynote: Simplicity Matters by Rich Hickey}} In every sphere
of a humans life, simplicity aligns perception with our mental
capacities. This paper should be read under the impression of
simplicity.

Derived from the ancient Latin word \textbf{simplex}, simple can be
understood as ``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points directly to the unidimensional aspect. While
complexity describes the multilayered und entangled nature of
conditions, simplicity empowers the human brain to reason about issues
in a straightforward manner. It certainly has some overlapping's with
easy, but while easy is more of a relative nature, simple can be laid
out as a objective manner - a goal software development should pursuit
and praise.

Software development is undoubtedly rich in complexity and full of
subtle pitfalls for the human brain. In a typical scenario, a piece of
software evolves over time in one or another \sout{opinionated}
direction. Layers of new abstractions wrestling with old legacy
abstractions and mutating objects will get out of control. Subtle bugs
start to creep in. Eventually the small piece of software ends up in a
highly complected monolith which will determine future design decisions
to a painful degree. Any company will be highly determined by the
current state trying to ``keep the lights on''.

On the other hand, any truly agile system architecture is laid out in a
fine-grained manner. As Rich Hickey argues, design decisions should be
made under the \textbf{impression of extending, substitution, moving,
combining and repurposing}. The ability to reason about the program at
any given time is crucial for future decisions and implementations.
Recalling again the unidimensional nature of simplicity.

@@

In an non-deterministic runtime environment, encapsulation and
modularization is a typical pattern to make complexity manageable and to
accommodate future uncertainty.{[}2, p. 1{]} Today many websites tend
towards becoming complex single-page applications (and therefore the
average JS payload is soaring too). Consequently, the demands to the
browser platform changed from a static page renderer to a
\textbf{complexe UI machine}. Under the current situation only
additional layers of abstraction can handle those complexity.

In the recent years many \textbf{frameworks}, libraries and
methodologies tried to tackle the global nature of the DOM by scoping
assets and design rules into maintainable components. While the DOM
can't be scoped, JS can. Therefore many of those frameworks, like
ReactJS, AngularJS or VueJS just to name a few, ditched the old rule of
separated HTML, CSS and JS in favor of an additional layer of abstracted
JS components (containing content, markup and styling). Quiet often
those frameworks mimic a MVC pattern on top of the browser engine which
is a reasonable simple design pattern to build graphical user
interfaces. This paper will apply the same pattern for building a
browsernative microservice.While frameworks are a valid approach to
build scalable web applications they remain highly opinionated, embody
inherent complexity themselves and can change and break over time.
Furthermore, they inflate the code with libraries. A standardized way
for creating complex UI interfaces painlessly requires new build-in
browser capabilities.

In 2013 thinkers, creators and browser vendors joined together to
propose \emph{The Extensible Web Manifesto}.\footnote{\href{https://extensiblewebmanifesto.org/}{The
  Extensible Web Manifesto}} The claim of the manifesto is to enhance
the current web platforms with new low-level capabilities. Those
features should empower creators of the web to write more declarative
code and therefore overcome known bottlenecks and artificial
abstractions. Four years later, the enhancement of JavaScript
leapfrogged and many new low-level APIs brought to life. With the new
APIs at hand a webdeveloper can create scoped and highly reusable
microservices without additional libraries even directly in the browser
console.

\emph{Disclaimer:} Currently, all new technologies proposed throughout
this paper can be tried and tested in Google Chrome, Opera and the
Safari Technology Preview. Usually the browser adoption time is short in
time an therefore full support of all standards is expected soon.
Meanwhile this standards can be used through \textbf{polyfills} on every
important legacy browser.

@@

Calling the case for \emph{Browsernative Microservices} brings up the
question about the concept of microservices in general. In fact the
concept of microservices has many facets, stretching beyond disciplines
and technical boundaries. It lacks a formal standardization but subsumes
certain ideas emergine from this pattern. As a primary source of truth
this paper relies on the work of Sam Newman, who has written a
comprehensive guide called \emph{Building Microservices} and the work of
Fowler and Lewis. The purpose of this section is to match those ideas
against the browser platform as targeted runtime.

In a nutshell a microservice is a small, autonomous service that works
together with other services seamlessly.{[}3, p. 2{]} or with the words
of Fowler and Lewis: ``\ldots{} the microservice architectural style is
an approach to developing a single application as a suite of small
services, each running in its own process and communicating with
lightweight mechanisms, often an HTTP resource API''{[}4{]}
Microservices incorporate many other ideas, like \emph{domain-driven
design} where we try to represent the real world in our code.{[}3, p.
2{]} Or making use of \emph{continuous delivery} for pushing software
rapidly through \emph{automated deployment} mechanisms in
production.{[}4{]} And, last but not least, microservices utilizes the
idea of small teams with a lot of product knowledge working mostly
autonomous on their very own service with their very own set of tools
and techniques.

@@

``A~\textbf{component}~is a unit of software that is independently
replaceable and upgradeable.''{[}4{]} Compared to libraries, a component
service offers multiple advantages. A library is only loosely coupled to
the implementation and therefore hard to track in functionality.
Changing a library often requires an unforeseen amount of time fixing
implementations. Therefore in the field we often see multiple versions
of libraries deployed simultaneous (like with JQuery) which adds a huge
amount of overhead on the page load. Eventually, dead code elimination
won't work in highly coupled libraries. Components instead incorporate a
much better evolution mechanism in the service contracts. Changing
functionality won't break other components. A component can
progressively enhanced which guarantees functionality throughout
different versions.

Compared to libraries componentized services offers a more explicit
interface.{[}4{]} While the functionality of a library needs
documentation, a component functionality is exposed via the components'
signature and therefore highly accessible. HTML elements are explicitly
controlled via their attributes and the new HTML service component are
no exception to that.

A microservice reinforces the \emph{Single Responsibility Principle}
defined by Robert C. Martin: ``Gather together those things that change
for the same reason and separate those things that change for different
reasons.''{[}5{]}

@@

\begin{quote}
``organizations which design systems \ldots{} are constrained to produce
designs which are copies of the communication~structures~of these
organizations''. {[}6{]}
\end{quote}

Microservice patterns are a product of real-world usage.{[}3, p. 1{]}
Instead of splitting teams along the technology stack (UI Experts
-\textgreater{} Middleware -\textgreater{} Database) a microservice
approach model teams around \textbf{business capabilities}.{[}4{]}
Consequently every team is capable of planning, designing, implementing,
testing and maintaining their very own microservice. Across the stack
every member gains high competence about the functionality of the
service.

By assigning service responsibility a team, the so called
\textbf{Definition of Done} (DoD) shifts from ``accomplishing projects''
to ``accomplishing products''. This new paradigm not only changes the
administrative overhead like budgeting or resource allocation. It
creates a kind of responsibility connection from the team to the
service. Expectedly, those teams are more motivated within their very
own service and exhibit a more sophisticated iteration time.{[}4{]}

For many companies working in the spheres of the internet, the
client-side is the most crucial area of business success. Todays
client-side includes many devices of differing sizes and
functionalities. As browser technologies, design guidelines and devices
change in a fast paced manner it makes absolutely sense to shift
responsibility towards the teams altogether. A \emph{browsernative
microservice} needs commitment over its whole lifecycle.

@@

Probably most microservices build for the browser are impure in
functionality. They aim for side-effects like changing the UI given to
input parameters or emitting browser events. Therefore a web developer
must design communication channels for those microservices keeping in
mind the nested nature of the DOM.

\emph{Smart endpoints and dumb pipes} is coined to the approach of
designing communication mostly decoupled and as cohesive as possible.
This essentially removes most of the error-prone middleware altogether.
In the analogy of the real world a message channel should look like
sending a letter: Only two endpoints (sender and receiver) connoted as
``smart'' and the letter itself looks mostly dumb (but unified) from the
outside. Only the receiver is able to process the content message.

Cited earlier in this paper microservices often rely on simple HTTP
request-response with resource API's and lightweight messaging as
communication protocols.{[}4{]} In the later parts of this paper the
reader will see a generic \emph{browsernative microservice} split in the
fashion of \textbf{MVC}. In the \textbf{VIEW} the microservice will rely
on browser events and in-memory calls while the \textbf{CONTROLLER} have
to deal with other resources over the fetch API.

\subsection{Decentralized Governance}\label{decentralized-governance}

\subsection{Decentralized Data
Management}\label{decentralized-data-management}

@@

The concept of microservices incooperates not only a technical
perspective.

Following this logic any company, whether it is web-related or not,
should be devided in units grouped around a destinct business service to
optimise the workflow. Fowler and Lewis outlines this approach as an
``alignment of business capabilities''{[}4{]} While this kind of
structure may be true for companies like Google or Amazon, there is a
vast majority of companies developing for the web which are grouped
around tasks.{[}7{]} A very common structure is formed by the technology
stack (UX Designers, Frontend- \& Backend Developers) or by separating
teams along the product lifecycle (development, testing, deployment).

>>>>

Computers can scale, humans can't. Ever since a program or complex
system made by humans has been constrained by humans mental
capabilities. Like in the analogy of juggling balls, our brain can just
``juggle'' a few things at a time. Rich Hickey, the inventor of the
programming language \emph{Clojure} gave an inspirational keynote on the
topic of \textbf{simplicity}.\footnote{\href{https://www.youtube.com/watch?v=rI8tNMsozo0\&t=46s}{Rails
  Conf 2012 Keynote: Simplicity Matters by Rich Hickey}} In every sphere
of a humans life, simplicity aligns perception with our mental
capacities.

Derived from the ancient Latin word \textbf{simplex}, simple can be
understood as ``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points directly to the unidimensional aspect. While
complexity describes the multilayered und entangled nature of
conditions, simplicity empowers the human brain to reason about issues
in a straightforward manner. It certainly has some overlapping's with
easy, but while easy is more of a relative nature, simple can be laid
out as a objective manner and therefore universally applicable.

Software development is undoubtedly rich in complexity and full of
subtle pitfalls for the human brain. In a typical scenario, a piece of
software evolves over time in one or another \sout{opinionated}
direction. Layers of new abstractions wrestling with old legacy
abstractions and mutation becomes untraceable. Subtle bugs start to
creep in. Eventually the small piece of software may end up in a highly
complected monolith which will determine future design decisions to a
painful degree. Future strategies of the company/organization will be
highly determined by the current state in the need of ``keeping the
lights on''.

On the other side of this dystopian example, the truly agile system
architecture is laid out in a fine-grained manner. Software becomes
pluggable, mutations traceable and modules interchangeable. As Rich
Hickey argues, design decisions should be made under the
\textbf{impression of extending, substitution, moving, combining and
repurposing}. The ability to reason about the program at any given time
is crucial for future decisions and implementations. Recalling again the
unidimensional nature of simplicity.

@@

In an non-deterministic runtime environment, encapsulation and
modularization is a typical pattern to make complexity manageable and
accommodate future uncertainty.{[}2, p. 1{]} Today many websites tend
towards becoming complex single-page applications (and therefore the
average JS payload is soaring too). Consequently, the demands to the
browser platform changed from a static page renderer to a
\textbf{complexe UI machine}. Under the current situation only
additional layers of abstraction can handle complexity.

In the recent years many \textbf{frameworks}, libraries and
methodologies approached the global nature of the DOM by scoping assets
and design rules into maintainable components. While the DOM can't be
scoped, JS can. Therefore many frameworks, like ReactJS, AngularJS or
VueJS just to name a few, ditched the old rule of separated HTML, CSS
and JS in favor of an additional layer of abstracted JS components
(containing content, markup and styling). Quiet often those frameworks
mimic a MVC pattern on top of the browser engine which is a reasonable
simple design pattern to build graphical user interfaces. While
frameworks are a valid approach to build scalable web applications they
remain highly opinionated, embody inherent complexity themselves and can
change and break over time. Furthermore, code inflation in front-end is
a crucial point for performance and third-party libraries are no
exception on that. A standardized way for creating complex UI interfaces
painlessly requires new build-in browser capabilities.

In 2013 thinkers, creators and browser vendors joined together to
propose \emph{The Extensible Web Manifesto}.\footnote{\href{https://extensiblewebmanifesto.org/}{The
  Extensible Web Manifesto}} The claim of the manifesto is to enhance
the current web platforms with new low-level capabilities. Those
features should empower creators of the web to write more declarative
code and therefore overcome known bottlenecks and artificial
abstractions. Four years later, the enhancement of JavaScript
leapfrogged and many new low-level APIs brought to life. With this new
APIs at hand a vivid web developer can create scoped and highly reusable
microservices without additional libraries even directly in the browser
console.

\emph{Disclaimer:} This paper introduces many new browser build-ins with
the focus on try and test. As the time of writing, examples can be tried
frictionless in the consoles of the latest versions of \textbf{Google
Chrome, Opera and Apple Safari}.\footnote{\href{http://jonrimmer.github.io/are-we-componentized-yet/}{Are
  we componentized yet?}} On Mozilla Firefox technologies work behind a
flag and Microsoft Edge implementation is unfortunately far behind. But
Browser implementation changes quickly and soon technology adoption
won't be an issue. Meanwhile new standards can be used through
\textbf{polyfills} even on legacy browsers.

@@

Calling the case for \emph{Browsernative Microservices} brings up the
question about the concept of microservices in general. In fact the
concept of microservices has many facets, stretching beyond disciplines
and technical boundaries. It lacks a formal standardization but subsumes
certain ideas emergine from this pattern. As a primary source of truth
this paper relies on the work of Sam Newman, who has written a
comprehensive guide called \emph{Building Microservices} and the work of
Fowler and Lewis. The purpose of this section is to match their ideas
against the browser platform as targeted runtime.

In a nutshell a microservice is a small, autonomous service that works
together with other services seamlessly.{[}3, p. 2{]} or with the words
of Fowler and Lewis: ``\ldots{} the microservice architectural style is
an approach to developing a single application as a suite of small
services, each running in its own process and communicating with
lightweight mechanisms, often an HTTP resource API.''{[}4{]}
Microservices incorporate many other ideas, like \emph{domain-driven
design} where it pursues the incorporation of the real world in our
code.{[}3, p. 2{]} Or making use of \emph{continuous delivery} for
pushing software rapidly through \emph{automated deployment} mechanisms
into production.{[}4{]} And, last but not least, microservices utilizes
the idea of small teams with a lot of product knowledge working mostly
autonomous on their very own service with their very own set of tools
and techniques.

@@

``A~\textbf{component}~is a unit of software that is independently
replaceable and upgradeable.'' {[}4{]} Components are the building
blocks of the microservice but in fact they even share a lot of
similarities just on a smaller scale. One of the similarity is the
\emph{loose coupling principle}: changing and deploying one service
shouldn't result in changing other parts of the system.{[}3, p. 30{]}.
Same goes for the component which should know as little as it needs
about the service to fulfil it's job.

Any reader of the paper coming from the ReactJS / Redux world this
concept of components might look familiar. Dan Abramov, the creator of
Redux, once defined a simple dichotomous pattern for creating UI
components. In his believe, there are \textbf{presentational components}
only related with the concern about \emph{how things look}. This
component literally doesn't know anything about the service in question
which makes the component highly flexible and reusable. They are
controlled solely from the outside, receiving data and dispatching
unbiased events on user interaction.{[}5{]}

On the other hand Dan Abramov outlines components which he calls
\textbf{containers}. A container component is concerned with \emph{how
things work}.{[}5{]} Containers acts as invisible wrappers around
presentational components. Their job is to fetch data from child nodes,
aggregating events, interacting with the model and push state back to
the presentational components. In contrast to their presentational
counterparts, containers can't life on its own without children.
Conceptually, they show another important principle of components and
microservices in general: \emph{high cohesion principle}.

Whether designing a microservice or it's components we want related
behavior sit together, and unrelated behavior to sit elsewhere.{[}3, p.
30{]} Code quality can be measured just by counting the places changes
in the code need to be made on changing behavior. In an arbitrary MVC
system a \texttt{button} might be inserted in the VIEW, the CONTROLLER
needs some adjustment and maybe the the MODEL, too. Three places for
adjustment is a reasonable easy task for the brain, everything more
violates the concept of simplicity.

Using libraries in web development is common sense. But compared to
libraries, a component service offers multiple advantages. A library is
only loosely coupled to the implementation and therefore hard to track
in functionality. Changing a library may result in an unforeseen amount
of time fixing implementations. It is not unusual to see websites
embodying different versions of the same library (like with JQuery).
Another issue with libraries is dead code elimination which means the
process of removing code that is never going to be executed. Newer build
tools for the web, like Webpack 2 or Rollup offer this feature which
relies heavily on the static structure on ES6 modules.{[}6{]} Contrary
libraries for the browsers are traditionally ``shipped'' as
\textbf{immediately-invoked functions}. Bootstrapping those closed
functions is much harder to archive as even dead code is actually
executed on load.

Web components are necessary static ES6 modules and any capable bundler
can choose which component is needed at runtime. Even more granular, a
good web component is composed of static JS blocks which can be strapped
away too.

A web component is self-contained which means it embodys all needed
functionality to get it's job done. Therefore it has a much better
evolution mechanism in the service contracts. Changing functionality
won't break other components. A component can progressively enhanced
which guarantees functionality throughout different versions.

Compared to libraries componentized services offers a more explicit
interface.{[}4{]} While the functionality of a library needs
documentation, a component functionality is exposed via the components'
signature which comes in the fashion of an HTML element. For example, a
button can consume attributes and becomes a primary buttons just by
assignment.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{<my-botton}\OtherTok{ isPrimary}\KeywordTok{>}\NormalTok{Primary}\KeywordTok{</my-botton>}
\end{Highlighting}
\end{Shaded}

A microservice reinforces the \emph{Single Responsibility Principle}
defined by Robert C. Martin: ``Gather together those things that change
for the same reason and separate those things that change for different
reasons.''{[}7{]}

@@

\begin{quote}
``organizations which design systems \ldots{} are constrained to produce
designs which are copies of the communication~structures~of these
organizations''. {[}8{]}
\end{quote}

Emphasizing the human factor in microservices is a key feature.
Microservices are a product of real-world usage.{[}3, p. 1{]} Instead of
splitting team structures along the technology stack (UI Experts
-\textgreater{} Middleware -\textgreater{} Database) a microservice
approach model teams around \textbf{business capabilities}.{[}4{]}
Consequently every team is capable of planning, designing, implementing,
testing and maintaining their very own microservice. Along the
technology stack every member gains high competence about the
functionality of the service.

Real-world domains tend to be complex and multifaceted. To unfold their
complexity, domains can subdivided into bounded contexts.{[}3, p. 31{]}
For example, dealing with customers is a business domain. Customers
again have different contexts depending on their demands. One context
can be sales, another context could be support. Every context makes
different assumptions about the underlying model. Each bounded context
draws an explicit interface where it decides what models to share with
other contexts.{[}3, p. 30{]} Each context can derived into multiple
microservices or, talking about \emph{browsernative microservices},
interfaces to the customers.

By assigning service responsibility to a team, the so called
\textbf{Definition of Done} (DoD) shifts from ``accomplishing projects''
to ``accomplishing products''. This new paradigm not only changes the
administrative overhead like budgeting or resource allocation. It
creates a kind of responsibility connection from the team to the
service. Expectedly, those teams are more motivated within their very
own service and exhibit a more sophisticated iteration time.{[}4{]}

For many companies working in the spheres of the internet the
client-side is highly important for their business. In fact, business
goals and capabilities can be derived from the front-end needs. The
state of the web is not only a story of numerous artifacts, it is also a
story of an highly fragmented market along devices, operating systems,
differing sizes and functionalities. Different devices again have
different assumptions about the technology stack. Splitting teams along
the stack results in an slow paced back and forth negotiation for every
change to be made. As browser technologies, design guidelines and
devices change in a fast paced manner it makes absolutely sense to shift
responsibility towards the teams altogether. A \emph{browsernative
microservice} embraces commitment over its whole lifecycle.

@@

Microservices for the browsers aim for side-effects like changing the UI
given to input parameters or emitting browser events. An isolated
microservice won't make too much sense after all. To ensure
communication between services, flexible yet powerful communication
channels must be established.

\emph{Smart endpoints and dumb pipes} is coined to the approach of
designing communication mostly decoupled and as cohesive as
possible.{[}4{]} In the analogy to the real world a message channel
should look like sending a letter: Two smart endpoints (sender and
receiver) and a mostly unified ``dumb'' letter format and channel. Each
endpoint owns their specific domain logic. In the browsernative context
we already emphasized the role of presentational components and
containers. The ``smartness'' of presentational components is altering
the UI and reacting on users input. The logic of a container follows the
idea of an filter in a Unix sense - receiving a request, applying logic
as appropriate and producing a response.{[}4{]}

Cited earlier in this paper microservices often rely on simple HTTP
request-response with resource API's and lightweight messaging as
communication protocols.{[}4{]} Those technologies highly accessible and
widely adopted and most probably usable for a browsernative service. In
the implementation part of this paper a reader will explore the
combination of REST protocols and a lightweight browser message bus in
action.

\subsection{Decentralization}\label{decentralization}

Decentralization is a meta concept of microservices. As written earlier
decentralization is part of the organizational structure by assigning
service responsibility solely to the team. Or by gluing independent
service bricks to a whole distributed system talking over the network
protocols.

This distributed nature allows teams to create their own technology
stack, tools and services designed in the spirit of language- and
platform independence - and share their knowledge with other
parties.{[}4{]} In the recent years many big companies like Facebook,
Google, Netflix and others followed that spirit and published their
ideas and implementations open source. The previously mentioned ReactJS
for example is a brainchild of Facebook. In fact, many tools and
techniques are byproduct of vital interaction of concrete domain
problems and their implementations.

The spirit of freedom can't be applied to \emph{browsernative
microservices} as the browser and its underlying DOM will be the
limitation factor. Still there are some decentralization aspects around
creating microservices for the browser worth to mention.

JS is the widely accepted language of the web. Nevertheless, a
microservice engineering team might choose another language for
different reasons. Transpiling languages to JS as target language isn't
exotic anymore. Languages like TypeScript, ClojureScript or PureScript
compile to JS even exclusively. Once web components hit a critical mass
there will be most likely some library support or foreign function
interfaces towards ES6 modules (which are mandatory for the new
specifications). With the rise of WebAssembly, a new low-level
programming language for the browser, the determination on JS will
hypothetical deteriorate and new quasi native languages for the web
might gain traction.

Another more real life decentralization aspect derives from the easiness
of deployment browsernative microservices which makes them perfectly for
open source. They are small, leightweight and can be tested live in the
browser without any additional tooling.
\href{https://webcomponents.org}{Webcomponents.org} for example is a
registry for many ready-to-use components where even Google shares a lot
of their material design elements.

Last bot not least, decentralized governance in a web native context
will be achieved just because of the encapsulation of the service.
Contrary to the HTML, CSS and JS split model, the new autonomous service
exhibits some sort of decentralization in creating the service.

@@

Following this logic any company, whether it is web-related or not,
should be devided in units grouped around a destinct business service to
optimise the workflow. Fowler and Lewis outlines this approach as an
``alignment of business capabilities''{[}4{]} While this kind of
structure may be true for companies like Google or Amazon, there is a
vast majority of companies developing for the web which are grouped
around tasks.{[}9{]} A very common structure is formed by the technology
stack (UX Designers, Frontend- \& Backend Developers) or by separating
teams along the product lifecycle (development, testing, deployment).
